---
title: "Darwin’s finches: Intraspecific morphological variation"
author: "Kirsten Hendricks"
date: "2025-03-18"
output: 
html_document : 
toc: true
toc_depth: 3
toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, 
               tidyr, 
               janitor,
               vegan,
               readr,
               magrittr,
               factoextra,
               ggpubr,
               ggplot2,
               rgl)
library(dynRB) 
library(MASS)
library(tibble)
library(ggplotify)
library(rmarkdown)
library(dynlm)
library(stats)
library(dplyr)
library(gtable)
library(glue)
library(knitr)
library(gt)
library(coin)
library(scales)
library(car)
library(moments)
library(effectsize)
library(fitdistrplus)



```

# Methods

### Sampling:

The first step is to fetch the data, and rename the species for readability.

```{r, echo=FALSE, 'fetching data + adding the sex variable'}
rm(list = ls()) #clear global environment
data("finch")   #fetching dataset
sex<-rep(c("F","M"),length(finch[,1])/2) #creating the new "sex" variable
fin2<-cbind(finch,sex) #adding the variable, as a new column, to the dataset
fin2 <- as.data.frame(fin2) %>% # Convert to data.frame for easier manipulation
  mutate(Species = case_when( 
        Species == "Geospiza_heliobates"    ~ "G. heliobates",      #Renaming species for readability
        
        Species == "Geospiza_prosthemelas_prosthemelas" ~ "G.p. prosthemelas",    #...
        
        Species == "Geospiza_fuliginosa_parvula"      ~ "G. f. parvula",    #...
        
        Species == "Geospiza_fortis_fortis"      ~ "G. f. fortis", #...........
    
         Species == "Geospiza_fortis_platyrhyncha"      ~ "G. f. platyrhyncha",#...
    

  TRUE ~ Species  # Keep other species  unchanged
  ))

df_fortis <- fin2 %>%
  filter(Species == "G. f. fortis")

df_platy <- fin2 %>%
  filter(Species == "G. f. platyrhyncha")

df_parvula <- fin2 %>%
  filter(Species == "G. f. parvula")

df_prosthem <- fin2 %>%
  filter(Species == "G.p. prosthemelas"
  )

df_heli <- fin2 %>%
  filter(Species == "G. heliobates"
  )


```

The data is then grouped by Species and the columns are labelled according to the traits being expressed by the each variable. The first, second and third statistical moments of each variable are then taken (those being the mean, standard deviation and variance respectively).

```{r echo=TRUE, message=TRUE, warning=TRUE, paged.print=TRUE}

finch_smry <- fin2 %>%
  group_by(Species) %>%
  dplyr::summarize(across(
    where(is.numeric),
    list(
      mean = ~ round(mean(., na.rm = TRUE), 3),
      sd = ~ round(sd(., na.rm = TRUE), 3),
      var = ~ round(var(., na.rm = TRUE), 3)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  pivot_longer(
    cols = -Species,
    names_to = c("Trait", ".value"),
    names_sep = "_"
  ) %>%
  mutate(Mean_SD = paste(mean, "±", sd)) %>%
  dplyr::select(-mean, -sd) %>%
  pivot_wider(
    id_cols = Species,
    names_from = Trait,
    values_from = c(Mean_SD, var),
    names_glue = "{Trait}_{.value}"
  ) %>%
  dplyr::select(  ##Reordering the columns such that the mean trait (± Std Dev) is directly adjascent to the trait varience. This means the readers' eyes do not need to scan back and forth across the whole table to find the variance of the "Upper Beak Length" (as an example), it is placed the row directly to the right of the mean and std dev row. 
    Species,
    BodyL_Mean_SD,
    BodyL_var,  
    WingL_Mean_SD, #Trait mean is ordered next to its matching trait variance
    WingL_var,
    TailL_Mean_SD,#................
    TailL_var,
    BeakW_Mean_SD,#..........................
    BeakW_var,
    BeakH_Mean_SD,#..................
    BeakH_var,
    LBeakL_Mean_SD,#................
    LBeakL_var,
    UBeakL_Mean_SD,#...................
    UBeakL_var,
    `N.UBkL_Mean_SD`,#...........................
    `N.UBkL_var`,
    TarsusL_Mean_SD,#...................
    TarsusL_var
  )

```

The following is a summary table denoting the different species of Finches, along with the means, ± the standard deviations and the variance of each variable.

```{r, echo=FALSE , results ='asis', 'summary statistical table'}

finch_smry%>%
gt()%>%
  cols_label(
    Species = "Species",
    BodyL_Mean_SD = "avg Body Length (± sd)",
    WingL_Mean_SD = "avg Wing Length (± sd)",
    TailL_Mean_SD = "avg Tail Length (± sd)",
    BeakW_Mean_SD = "avg Beak Width (± sd)",
    BeakH_Mean_SD = "avg Beak Height (± sd)",
    LBeakL_Mean_SD ="avg L.Beak Length (± sd)",
    UBeakL_Mean_SD = "avg U. Beak Length (± sd)",
    `N.UBkL_Mean_SD` = "N to U. Beak Length (± sd)",
    TarsusL_Mean_SD = "avg Tarsus Length (± sd)",
    BodyL_var = "BL var",
    WingL_var = "WL var",
    TailL_var = "TL var",
    BeakW_var = "BW var",
    BeakH_var = "BH var",
    LBeakL_var = "LBL var",
    UBeakL_var = "UBL var",
    `N.UBkL_var` = "NUBKL var",
    TarsusL_var= "TarsusL var"
  ) %>%
  data_color(
    columns = Species,
    rows = everything(),
    method = "factor",
    palette =  c(
        "G. f. fortis" = "purple",
        "G. f. parvula" = "green",
        "G. f. platyrhyncha" = "cyan",
        "G. heliobates" = "red",
        "G.p. prosthemelas" = "grey"),
    alpha = 0.8
  )

```

### Data Analysis

MANOVA:

```{r pressure, echo=TRUE}

##Creating a List to assign Species Colours Once and for all

species_colors <- c(
  "G. heliobates" = "red",
  "G.p. prosthemelas" = "grey",
  "G. f. parvula" = "green",
  "G. f. fortis" = "purple",
  "G. f. platyrhyncha" = "cyan"
)

##Box Plots for comparing Body Lengths of the species

ggplot2::ggplot(data= fin2, aes(x = Species, y = BodyL, fill = Species)) +
  geom_boxplot() +
  scale_fill_manual(values = species_colors) +
  xlab("Species") +
  ylab("Body Length (mm)") +
  theme_classic() +
  #facet_wrap(~Species) +
  guides(fill = guide_legend(title = "Species"))

# Using data.frame 'fin2' and the morphological variables therein:

# we perform the MANOVA on the data, grouped by species

depvar <- cbind(fin2$BodyL, fin2$WingL, fin2$TailL, fin2$BeakW, fin2$BeakH, fin2$LBeakL, fin2$UBeakL, fin2$N.UBkL, fin2$TarsusL)

manova_model <- manova(depvar ~ Species, data = fin2)
print(manova_model)


# print and summarise the results


summary(manova_model) 

eta_squared(manova_model)
```

PCA

```{r, echo = TRUE}
      
      # Let's first check whether there are any NA values
      any(is.na(fin2))
      FinchData = na.omit(fin2)
      
      
      # remove non-numeric columns from the dataframe (Species and sex)
      FinchDataclean = FinchData %>%
        dplyr::select(where(is.numeric))
      
      # scaling the data
      pca_fit <- FinchDataclean %>% 
        prcomp(scale = TRUE) 
      
      # PCA summary results
      summary(pca_fit)
   
      
      #loadings
      pca_loadings = pca_fit$rotation %>%
        as.data.frame() %>%
        dplyr::select(c(PC1, PC2))
      
      
      # Collecting the PC scores
      pca_fit_df = pca_fit$x %>%
        as.data.frame() %>%
        dplyr::select(c(PC1, PC2)) #To focus on the first 2 principle components
      
    
      # Add "Species" column to the end of pca_fit_df dataframe
      pca_fit_df$Species = FinchData$Species
      
      # change the alpha value to add transparency to the colours
      pca_plot = pca_fit_df %>%
        ggplot(aes(PC1, PC2, color = Species)) + 
        geom_point(size = 1.5, alpha = 0.5) +
        scale_color_manual(
          values = species_colors
        ) +
        xlab("PC1 ") +
        ylab("PC2 ") +
        labs(color = "Species",
             title = "PC1 and PC2 for Darwin's Finches ") +
        # add lines through the origin 
        geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
        geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
        theme_classic()
      
      pca_plot
      
```

LDA:

```{r, echo = TRUE}

# Select the relevant columns for LDA
lda_data <- fin2[, c("Species", "BodyL", "WingL", "TailL", "BeakW", "BeakH", "LBeakL", "UBeakL", "N.UBkL", "TarsusL")]

# Perform LDA
lda_result <- lda(Species ~ ., data = lda_data)

lda_df <- lda_result$scaling %>% 
  as.data.frame()%>%
  dplyr::select(c(LD1, LD2, LD3, LD4)) 

  gt(lda_df,
  rowid_to_column(lda_df)%>% head(),
  rowname_col = "Traits",
  caption = ""
  
  

)






```

```{r, echo = TRUE}


# Log-transform all variables for "D. F. Fortis" bar sex 

df_fortis$logBodyL_f <- log(df_fortis$BodyL)
df_fortis$logWingL_f <- log(df_fortis$WingL)
df_fortis$logBeakH_f  <- log(df_fortis$BeakH)

#//////////////////////////////////////////////////////////////////////////////////////////////

# Fit linear models to the log-transformed data in a pairwise manner per species



FortBody_Wing_allom <- lm(logBodyL_f   ~ logWingL_f, data = df_fortis)
summary(FortBody_Wing_allom)

ggplot( data = df_fortis, aes(x= logBodyL_f, y= logWingL_f)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Wing Length") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.2, colour = "purple", linewidth = 0.3, se = TRUE)

  


FortBeakH_Body_allom <- lm(logBodyL_f ~ logBeakH_f, data = df_fortis)
summary(FortBeakH_Body_allom)

ggplot( data = df_fortis, aes(x= logBodyL_f, y=logBeakH_f)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Beak height") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "purple", linewidth = 0.3, se = TRUE)

  


##/////////////////////////////////////////////////////////////////////////////////////////////


# Examine the slopes of the regression lines





slope_BodW_fort <- coef(summary(FortBody_Wing_allom))[2, 1]

print(slope_BodW_fort)

p_value_BodW_fort <- coef(summary(FortBody_Wing_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Wing Length) for G. f. fortis:", slope_BodW_fort, "\n")

cat("P-value for the slope:", p_value_BodW_fort, "\n")

#########

slope_BodBkH_fort <- coef(summary(FortBeakH_Body_allom))[2, 1]

print(slope_BodBkH_fort)

p_value_BodBkH_fort <- coef(summary(FortBeakH_Body_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Beak Height) for G. f. fortis:", slope_BodBkH_fort, "\n")

cat("P-value for the slope:", p_value_BodBkH_fort, "\n")


##########

fort_slopes <- cbind(slope_BodW_fort, slope_BodBkH_fort)

# check sample sizes with dplyr so r knows which package to source the 'count' function for/ prevents coersion errors:
fin2 %>%
  dplyr::count(Species)

#Perform a t-test to check scaling = allometric ( if slope != 1, then body features scale allometrically, else the scaling is isometric)

t_test_fort <- t.test(fort_slopes, mu = 1)

print(t_test_fort) ##t-test result shows that the true mean is not equal to 1


#########################################################################


# log transform variables for "G. F. Parvula"

df_parvula$logBodyL_pa <- log(df_parvula$BodyL)
df_parvula$logWingL_pa <- log(df_parvula$WingL)
df_parvula$logBeakH_pa  <- log(df_parvula$BeakH)

# Fit linear models to the log-transformed data in a pairwise manner per species



Parv_Body_Wing_allom <- lm(logBodyL_pa   ~ logWingL_pa, data = df_parvula)
summary(Parv_Body_Wing_allom)

ggplot( data = df_parvula, aes(x= logBodyL_pa, y=logWingL_pa)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Wing Length") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "green", linewidth = 0.3, se = TRUE)

Parv_BeakH_Body_allom <- lm(logBeakH_pa ~ logBodyL_pa, data = df_parvula)
summary(Parv_BeakH_Body_allom)

ggplot( data = df_parvula, aes(x= logBodyL_pa, y=logBeakH_pa)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Beak height") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "green", linewidth = 0.3, se = TRUE)


##/////////////////////////////////////////////////////////////////////////////////////////////


# Examine the slopes of the regression lines


slope_BodW_parv <- coef(summary(Parv_Body_Wing_allom))[2, 1]

print(slope_BodW_fort)

p_value_BodW_parv <- coef(summary(Parv_Body_Wing_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Wing Length) for  G. f. parvula:", slope_BodW_parv, "\n")

cat("P-value for the slope:", p_value_BodW_parv, "\n")

#########

slope_BodBkH_parv <- coef(summary(Parv_BeakH_Body_allom))[2, 1]

print(slope_BodBkH_parv)

p_value_BodBkH_parv <- coef(summary(Parv_BeakH_Body_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Beak Height) for 	G. f. parvula:", slope_BodBkH_parv, "\n")

cat("P-value for the slope:", p_value_BodBkH_parv, "\n")


##########

parv_slopes <- cbind(slope_BodW_parv, slope_BodBkH_parv)

# check sample sizes with dplyr so r knows which package to source the 'count' function for/ prevents coersion errors:
fin2 %>%
  dplyr::count(Species)

#Perform a t-test to check scaling = allometric ( if slope != 1, then body features scale allometrically, else the scaling is isometric)

t_test_parv <- t.test(parv_slopes, mu = 1)

print(t_test_parv) ##t-test result shows that the true mean is not equal to 1


#########################################################################

#Log transform data for G. f. platyrhyncha

df_platy$logBodyL_pl <- log(df_platy$BodyL)
df_platy$logWingL_pl <- log(df_platy$WingL)
df_platy$logBeakH_pl  <- log(df_platy$BeakH)

# Fit linear models to the log-transformed data in a pairwise manner per species



Platy_Body_Wing_allom <- lm(logBodyL_pl   ~ logWingL_pl, data = df_platy)
summary(Platy_Body_Wing_allom)

ggplot( data = df_platy, aes(x= logBodyL_pl, y=logWingL_pl)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Wing Length") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "cyan", linewidth = 0.3, se = TRUE)

Platy_BeakH_Body_allom <- lm(logBodyL_pl ~ logBeakH_pl, data = df_platy)
summary(Platy_BeakH_Body_allom)


ggplot( data = df_platy, aes(x= logBodyL_pl, y=logBeakH_pl)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Beak height") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "cyan", linewidth = 0.3, se = TRUE)


##/////////////////////////////////////////////////////////////////////////////////////////////


# Examine the slopes of the regression lines





slope_BodW_pl <- coef(summary(Platy_Body_Wing_allom))[2, 1]

print(slope_BodW_pl)

p_value_BodW_pl <- coef(summary(Platy_Body_Wing_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Wing Length) for G. f. platyrhyncha:", slope_BodW_pl, "\n")

cat("P-value for the slope:", p_value_BodW_pl, "\n")

#########

slope_BodBkH_pl <- coef(summary(Platy_BeakH_Body_allom))[2, 1]

print(slope_BodBkH_pl)

p_value_BodBkH_pl <- coef(summary(Platy_BeakH_Body_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Beak Height) for G. f. platyrhyncha:", slope_BodBkH_pl, "\n")

cat("P-value for the slope:", p_value_BodBkH_pl, "\n")


##########

platy_slopes <- cbind(slope_BodW_pl, slope_BodBkH_pl)

# check sample sizes with dplyr so r knows which package to source the 'count' function for/ prevents coersion errors:
fin2 %>%
  dplyr::count(Species)

#Perform a t-test to check scaling = allometric ( if slope != 1, then body features scale allometrically, else the scaling is isometric)

t_test_pl <- t.test(platy_slopes, mu = 1)

print(t_test_pl) ##t-test result shows that the true mean is not equal to 1



#########################################################################

#Log Transform the data for G. heliobates

df_heli$logBodyL_h <- log(df_heli$BodyL)
df_heli$logWingL_h <- log(df_heli$WingL)
df_heli$logBeakH_h  <- log(df_heli$BeakH)

# Fit linear models to the log-transformed data in a pairwise manner per species


Heli_Body_Wing_allom <- lm(logBodyL_h   ~ logWingL_h, data = df_heli)
summary(Heli_Body_Wing_allom)

ggplot( data = df_heli, aes(x= logBodyL_h, y=logWingL_h)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Wing Length") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "red", linewidth = 0.3, se = TRUE)

Heli_BeakH_Body_allom <- lm(logBodyL_h ~ logBeakH_h, data = df_heli)
summary(Heli_BeakH_Body_allom)


ggplot( data = df_heli, aes(x= logBodyL_h, y=logBeakH_h)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Beak height") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "red", linewidth = 0.3, se = TRUE)



##/////////////////////////////////////////////////////////////////////////////////////////////


# Examine the slopes of the regression lines



slope_BodW_he <- coef(summary(Heli_Body_Wing_allom))[2, 1]

print(slope_BodW_he)

p_value_BodW_he <- coef(summary(Heli_Body_Wing_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Wing Length) for G. heliobates:", slope_BodW_he, "\n")

cat("P-value for the slope:", p_value_BodW_he, "\n")

#########

slope_BodBkH_he <- coef(summary(Heli_BeakH_Body_allom))[2, 1]

print(slope_BodBkH_he)

p_value_BodBkH_he <- coef(summary(Heli_BeakH_Body_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Beak Height) for G. heliobates:", slope_BodBkH_he, "\n")

cat("P-value for the slope:", p_value_BodBkH_he, "\n")


##########

heli_slopes <- cbind(slope_BodW_he, slope_BodBkH_he)

# check sample sizes with dplyr so r knows which package to source the 'count' function for/ prevents coersion errors:
fin2 %>%
  dplyr::count(Species)

#Perform a t-test to check scaling = allometric ( if slope != 1, then body features scale allometrically, else the scaling is isometric)

t_test_heli <- t.test(heli_slopes, mu = 1)

print(t_test_heli) ##t-test result shows that the true mean is not equal to 1



#########################################################################

#Log Transform data for G.p. prosthemelas

df_prosthem$logBodyL_pr <- log(df_prosthem$BodyL)
df_prosthem$logWingL_pr <- log(df_prosthem$WingL)
df_prosthem$logBeakH_pr <- log(df_prosthem$BeakH)

# Fit linear models to the log-transformed data to compare relelationship


Pros_Body_Wing_allom <- lm(logBodyL_pr   ~ logWingL_pr, data = df_prosthem)
summary(Pros_Body_Wing_allom)

ggplot( data = df_prosthem, aes(x= logBodyL_pr, y=logWingL_pr)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Wing Length") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "grey", linewidth = 0.3, se = TRUE)

Pros_BeakH_Body_allom <- lm(logBodyL_pr ~ logBeakH_pr, data = df_prosthem)
summary(Pros_BeakH_Body_allom)



ggplot( data = df_prosthem, aes(x= logBodyL_pr, y=logBeakH_pr)) +
  geom_point() +
   xlab("Log Body Length") +
   ylab("Log Beak height") +
   theme_classic() +
   stat_smooth(method = "lm", geom ="smooth", 
   alpha = 0.4, colour = "grey", linewidth = 0.3, se = TRUE)


##/////////////////////////////////////////////////////////////////////////////////////////////


# Examine the slopes of the regression lines



slope_BodW_pr <- coef(summary(Pros_Body_Wing_allom))[2, 1]

print(slope_BodW_fort)

p_value_BodW_pr <- coef(summary(Pros_Body_Wing_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Wing Length) for G.p. prosthemelas:", slope_BodW_pr, "\n")

cat("P-value for the slope:", p_value_BodW_pr, "\n")

#########

slope_BodBkH_pr <- coef(summary(Pros_BeakH_Body_allom))[2, 1]

print(slope_BodBkH_pr)

p_value_BodBkH_pr <- coef(summary(Pros_BeakH_Body_allom))[2, 4]

cat("Slope of log(Body Length) vs log(Beak Height) for G.p. prosthemelas:", slope_BodBkH_pr, "\n")

cat("P-value for the slope:", p_value_BodBkH_pr, "\n")


##########

Pr_slopes <- cbind(slope_BodW_pr, slope_BodBkH_pr)

# check sample sizes with dplyr so r knows which package to source the 'count' function for/ prevents coersion errors:
fin2 %>%
  dplyr::count(Species)

#Perform a t-test to check scaling = allometric ( if slope != 1, then body features scale allometrically, else the scaling is isometric)

t_test_pr <- t.test(Pr_slopes, mu = 1)

print(t_test_pr) ##t-test result shows that the true mean is not equal to 1




#########################################################################



#########################################################################



```

```{r, echo = TRUE}


ggplot(df_fortis, aes(x = BeakH)) +
  geom_histogram(bins = 10, fill = "purple", color = "black") +
  xlab("Beak Height (mm)") +
  ylab("Frequency") +
  ggtitle("Frequency Distribution of Beak Height in Geospiza f. fortis")

descdist(df_fortis$BeakH, discrete = FALSE)

fit.weibull <- fitdist(df_fortis$BeakH, "weibull")
fit.norm <- fitdist(df_fortis$BeakH, "norm")

#checking fit for normal distriibution

plot(fit.norm) ##Looks more like Normal dist, based on Q-Q plot 

#Weibull Distribution fit-check

plot(fit.weibull) ##Q-Q plot has an outlier on the low end likely to be a worse fit
```

```{r, echo = TRUE}


# Calculate skewness
skewness_value <- skewness(df_fortis$BeakH)
cat("Skewness of Beak Height in G. f. fortis:", skewness_value, "\n")

# Calculate kurtosis 
kurtosis_value <- kurtosis(df_fortis$BeakH)
cat("Kurtosis of Beak Height in G. f. fortis:", kurtosis_value, "\n")

```

```{r, echo = TRUE}
# Separate data by sex
bodyL_female <- df_fortis %>% filter(sex == "F") %>% pull(BodyL)
bodyL_male <- df_fortis %>% filter(sex == "M") %>% pull(BodyL)


# Visual inspection: Histograms
par(mfrow = c(1, 2))
hist(bodyL_female, main = "Body Length (Female)", col = "purple", xlab = "mm")
hist(bodyL_male, main = "Body Length (Male)", col = "purple", xlab = "mm")
par(mfrow = c(1, 1))

# Visual inspection: Density plots
plot(density(bodyL_female), main = "Body Length (Female)")
plot(density(bodyL_male), main = "Body Length (Male)")

# Statistical test for normality (Shapiro-Wilk)
shapiro.test(bodyL_female)
shapiro.test(bodyL_male)

# Levene's test


leveneTest(BodyL ~ sex, data = df_fortis)

```

```{r, echo = TRUE}
# Separate data by sex
tailL_female <- df_fortis %>% filter(sex == "F") %>% pull(TailL)
tailL_male <- df_fortis %>% filter(sex == "M") %>% pull(TailL)

# Visual inspection: Histograms
par(mfrow = c(1, 2))
hist(tailL_female, main = "Tail Length (Female)", col = "purple", xlab = "mm")
hist(tailL_male, main = "Tail Length (Male)", col= "purple", xlab = "mm")
par(mfrow = c(1, 1))

# Visual inspection: Density plots
plot(density(tailL_female), main = "Tail Length (Female)")
plot(density(tailL_male), main = "Tail Length (Male)")

# Statistical test for normality (Shapiro-Wilk)
shapiro.test(tailL_female)
shapiro.test(tailL_male)

# Statistical test for equality of variances (Levene's test)
leveneTest(TailL ~ sex, data = df_fortis)
```

## References

Ben-Shachar M, Lüdecke D, Makowski D (2020). effectsize: Estimation of Effect Size Indices and Standardized Parameters. Journal of Open Source Software, 5(56), 2815. doi: 10.21105/joss.02815

Hester J, Bryan J (2024). *glue: Interpreted String Literals*. R package version 1.8.0, <https://CRAN.R-project.org/package=glue>.

Komsta L, Novomestky F (2022). *moments: Moments, Cumulants, Skewness, Kurtosis and Related Tests*. R package version 0.14.1, <https://CRAN.R-project.org/package=moments>.

Marie Laure Delignette-Muller, Christophe Dutang (2015). fitdistrplus: An R Package for Fitting Distributions. Journal of Statistical Software, 64(4), 1-34. DOI 10.18637/jss.v064.i04.

```         
Rinker, T. W. & Kurkiewicz, D. (2017). pacman: Package Management for R.
```

version 0.5.0. Buffalo, New York. <http://github.com/trinker/pacman>

```         
Snodgrass R and Heller E (1904) Papers from the Hopkins-Stanford Galapagos            Expedition, 1898-99. XVI. Birds. Proceedings of the Washington Academy of Sciences 5: 231-372.

Venables, W. N. & Ripley, B. D. (2002) Modern Applied Statistics with S.
```

Fourth Edition. Springer, New York. ISBN 0-387-95457-0

```         
 Wickham H, François R, Henry L, Müller K, Vaughan D (2023). _dplyr: A
```

Grammar of Data Manipulation\_. R package version 1.1.4, <https://CRAN.R-project.org/package=dplyr>.

Xie Y (2025). *knitr: A General-Purpose Package for Dynamic Report Generation in R*. R package version 1.50, <https://yihui.org/knitr/>.
